Optimizing Abstract Abstract Machines

### 2013-07-02 paper/paper.tex
Uncomment some acks.
Tightened some wording and moved some less-important comments to footnotes.

\documentclass[preprint,onecolumn,9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{balance}
\usepackage{calc}

@@

\conferenceinfo{WXYZ '05}{date, City.}
\copyrightyear{2005}
\copyrightdata{[to be supplied]}

@@

Program analysis provides sound predictive models of program behavior, but in order for such models to be effective, they must be efficiently computable and correct.
%
Past approaches to designing program analysis have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
%
The \emph{abstracting abstract machines} (AAM) approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic} to deriving program analyses provides an alternative: a systematic way of transforming a programming language semantics in the form of an abstract machine into a family of abstract interpreters.
%
It thus reduces the burden of constructing and verifying the soundness of an abstract interpreter.
%

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{
Example state graphs for Earl et. al. program.
%
  Gray states follow variable references, {\tt ev} states are black, and all others are white.
%
 Part (a) shows the result of the baseline analyzer.
%
  It has long ``corridor'' transitions and ``diamond'' subgraphs that fan-out from nondeterminism and fan-in from joins.
%
  Part (b) shows the result of performing nondeterminism lazily and thus avoids many of the diamond subgraphs.
%
  Part (c) shows the result of abstract compilation that removes interpretive overhead in the form of intermediate states, thus minimizing the corridor transitions.
%
  The end result is a more compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

The semantics are defined in terms of a machine model.
%
The machine components are defined in figure~\ref{fig:domains};
%
figure~\ref{fig:aam} defines the transition relation (unmentioned components stay the same).
%
The evaluation of a program is defined as its set of traces that arise from iterating the machine transition relation.
%
The $\traces$ function produces the set of all proofs of reachability for any state $\mstate$ from the injection of program $\mexp$ (from which one could extract a string of states).
%
 The machine is a very slight variation on a standard abstract machine for ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.
%
It can be systematically derived from a definitional interpreter through a continuation-passing style transformation and defunctionalization, or from a structural operational semantics using the refocusing construction of Danvy and Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
%%

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction where arithmetic operations return an abstract top element
$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
$\sNum$.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is a sound abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  The former is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; the latter is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.

\paragraph{Uniform \(k\)-CFA} To characterize uniform \(k\)-CFA, set the allocation
strategy as follows, for a fixed constant \(k\):

{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\epsilon} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as
the application of a widening operator\footnote{Technically, we would
  have to copy the value of the global store to all states being
  stepped to fit the formal definition of a widening, but this
  representation is order-isomorphic to that.} during the fixed-point
iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it installs this joined store into current
state.

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
{\small
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
%%\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}}
%
The new store is computed as the least upper bound of all
stores after stepping in order to have only one, and keep the
semantics sound.

@@

The final approximation we make to get to our baseline is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the \kfnalone
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected). We also need to allocate an address
for the argument position in an \apalone state.

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially (${\mathcal O}(n^n)$). We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfnalone$ continuation. The continuation has to contain this address
to remember where to retrieve values from in the store.

The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
\text{ where } \maddr &= \alloc(\mstate) \\
               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}

@@

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin \mseen}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  \mseen' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we
have seen the first state, and that we need to process the first
state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

Notice that now $\mseen$ has several copies of the abstract store in it. As
it is, this semantics is much less efficient (but still more precise)
than the previously proposed semantics because membership checks have
to compare entire stores. Checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

And, there is a better way. Shivers' original work on $k$-CFA was
susceptible to the same problem, and he suggested three complementary
optimizations: (1) make the store global; (2) update the store
imperatively; and (3) associate every change in the store with a
version number -- its timestamp. Then, put timestamps in states
where previously there were stores. Given two states, the analysis can
now compare their stores just by comparing their timestamps -- a
constant-time operation.

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in
the following way:
{\small
\begin{align*}
(\mseentime, F, \msto,\Sigma,t) &\mathrel{\damachstep} (\mseentime \cup \mseentime', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
 \mseentime' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}}

Here $\appendall$ combines change logs across all non-deterministic
steps for a state to later be replayed. The order the combination
happens in doesn't matter, because join is associative and
commutative.

\begin{lemma}
$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff
$\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

\begin{lemma}[$\changep$ means change]
Let $\replay(\msdiff,\msto) = \ttuple{\msto'}{\changep}$. $\msto' \neq \msto$ iff $\changep$.
\end{lemma}
By induction on $\msdiff$.

\begin{theorem}
$\damachstep$ is a complete abstraction of $\widehat{\machstep}^T$.
\end{theorem}
Follows from previous lemma and that join is associative and commutative.

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions.
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.2]{fanoutdot2}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if}, a called procedure, or a numerical primitive application).
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.2]{lazydot}
\end{center}

This shift does not change the concrete semantics of the language to
be lazy.  Rather, it abstracts over transitions that the original
non-deterministic semantics steps through.
%
We say the abstraction is \emph{lazy} because it delays splitting on
the values in an address until they are \emph{needed} in the
semantics. It does not change the execution order that leads to the
values that are stored in the address.

@@

\renewcommand{\ext}{\mathit{ext}}
{\small
\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\msto,\mval) &= \setof{\mval}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\mkaddr),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.

\spchoice{
\noindent
We have two choices for how to implement lazy non-determinism.

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step, since store widening
leads to stepping several states in one big ``step.'' With laziness,
the reference will mean the original binding(s) of the variable or the
new binding, because the actual store lookup is delayed one step
(i.e. laziness is administrative).

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by storing the result of the lookup in a value
representing a delayed nondeterministic choice. This is a more common
choice in 0CFA implementations we have seen, but it interferes with
the next optimization (discussion in next section).

\paragraph{Our choice: option 1}
The conditions for precision loss happen too rarely to warrant the
added complexity to the next section's semantics to avoid such
losses. Without abstract compilation, option 2 is the better
choice.}{}

@@

Thus far, we have made our optimizations in a purely functional
manner. For the final push for performance, we need to dip into the
imperative. In this section, we show an alternative representation of
the store and seen set that are more space-efficient and are amenable
to destructive updates by adhering to a history for each address.

>>>>

\documentclass[9pt]{sigplanconf} %{onecol}
\usepackage{alltt,mathpartir}
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{url}
\usepackage{graphicx}
\usepackage{balance}
\usepackage{calc}

@@

\conferenceinfo{ICFP '13}{September 25--27, Boston.}
\copyrightyear{2013}
\copyrightdata{[to be supplied]}

@@

Program analysis provides sound predictive models of program behavior, but in order for such models to be effective, they must be efficiently computable and correct.
%
Past approaches to designing program analyses have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
%
The \emph{abstracting abstract machines} (AAM) approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic} to deriving program analyses provides an alternative: a systematic way of transforming a programming language semantics in the form of an abstract machine into a family of abstract interpreters.
%
It thus reduces the burden of constructing and verifying the soundness of an abstract interpreter.
%

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{
Example state graphs for Earl et. al. program.
%
  Gray states follow variable references, {\tt ev} states are black, and all others are white.
%
 Part (a) shows the baseline analyzer result.
%
  It has long ``corridor'' transitions and ``diamond'' subgraphs that fan-out from nondeterminism and fan-in from joins.
%
  Part (b) shows the result of performing nondeterminism lazily and thus avoids many of the diamond subgraphs.
%
  Part (c) shows the result of abstract compilation that removes interpretive overhead in the form of intermediate states, thus minimizing the corridor transitions.
%
  The end result is a more compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

The semantics is defined in terms of a machine model.
%
The machine components are defined in figure~\ref{fig:domains};
%
figure~\ref{fig:aam} defines the transition relation (unmentioned components stay the same).
%
The evaluation of a program is defined as its set of traces that arise from iterating the machine transition relation.
%
The $\traces$ function produces the set of all proofs of reachability for any state $\mstate$ from the injection of program $\mexp$ (from which one could extract a string of states).
%
 The machine is a very slight variation on a standard abstract machine for ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.
%
It can be systematically derived from a definitional interpreter through a continuation-passing style transformation and defunctionalization, or from a structural operational semantics using the refocusing construction of Danvy and Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
%%

@@

\paragraph{Abstract interpretation}
%
To characterize abstract interpretation, set the implicit parameters just as above, but drop the $\maddr \not\in \msto$ condition.
%
The $\interpdelta$ relation takes some care to not make the analysis run forever; a simple instantiation is a flat abstraction where arithmetic operations return an abstract top element $\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on $\sNum$.
%
This family of interpreters is also non-deterministic in choices of addresses, but it is free to choose addresses that are already in use.
%
Consequently, the machines may be non-deterministic when multiple values reside in a store location.
%%

%%
It is important to recognize from this definition that \emph{any} allocation strategy is a sound abstract interpretation~\cite{dvanhorn:Might2009Posteriori}.
%
In particular, concrete interpretation is a kind of abstract interpretation.
%
So is an interpretation that allocates a single cell into which all bindings and continuations are stored.
%
The former is an abstract interpretation with uncomputable reachability and gives only the ground truth of a program's behavior; the latter is an abstract interpretation that is easy to compute but gives little information.
%
Useful program analyses lay somewhere in between and can be characterized by their choice of address representation and allocation strategy.
%
Uniform \(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.
%%

%%
\paragraph{Uniform \(k\)-CFA}
%
To characterize uniform \(k\)-CFA, set the allocation strategy as follows, for a fixed constant \(k\):
%%

%%
{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\epsilon} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

A common technique to accelerate convergence in flow analyses is to share a common, global store.
%
Formally, we can cast this optimization as a second abstraction or as the application of a widening operator
%
\footnote{Technically, we would have to copy the value of the global store to all states being stepped to fit the formal definition of a widening, but this representation is order-isomorphic to that.}
%
during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the program.
%
Thus, complexity drops from intractable exponentiality to a merely daunting polynomial.
%%

%%
Since we can cast this optimization as a widening, there is no need to change the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join them together.
%
Then, before each transition, it installs this joined store into current state.
%%

%%
To describe this process, AAM defined a transformation of the reduction relation so that it operates on a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
{\small
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto'), \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn &: \Context \times \Store \to \State \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
%%\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}}
%
To retain soundness, this store grows monotonically as the least upper bound of all occurring stores.
%
%The new store is computed as the least upper bound of all stores after stepping in order to have only one, and keep the semantics sound.

@@

The final approximation we make to get to our baseline is to store-allocate all values that appear, so that any non-machine state that contains a value instead contains an address to a value.
%
The AAM approach stops at the previous optimization.
%
However, the \kfnalone continuation stores a value, and this makes the space of continuations quadratic rather than linear in the size of the program, for a monovariant analysis like 0-CFA.
%
Having the space of continuations grow linearly with the size of the program will drop the overall complexity to cubic (as expected).
%
We also need to allocate an address for the argument position in an \apalone state.
%%

%%
To achieve this linearity for continuations, we allocate an address for the value position when we create the continuation.
%
This address and the tail address are both determined by the label of the application point, so the space becomes linear and the overall complexity drops to cubic.
%
This is a critical abstraction in languages with $n$-ary functions, since otherwise the continuation space grows super-exponentially (${\mathcal O}(n^n)$).
%
We extend the semantics to additionally allocate an address for the function value when creating the $\kfnalone$ continuation.
%
The continuation has to contain this address to remember where to retrieve values from in the store.
%%

%%
The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
\text{ where } \maddr &= \alloc(\mstate) \\
               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}

@@

\begin{equation*}
\msto' = \bigsqcup\; \setof{\msto \mid (\_,\msto) \in I}
\end{equation*}
%
The next frontier is exactly the states that we found from stepping
the last frontier, but have not seen before. They must be
\emph{states}, so we pair the contexts with the next store:
%
\begin{equation*}
F' = \setof{c \mid (c,\_) \in I, (c,\msto') \notin \mseen}
\end{equation*}
%
Finally, we add what we know we had not yet seen to the seen set:
\begin{equation*}
  \mseen' = \setof{(c,\msto') \mid c \in F'}
\end{equation*}
%
To inject a program $e$ into this machine, we start off knowing we have seen the first state, and that we need to process the first state:
\begin{align*}
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}
%%

%%
Notice that now $\mseen$ has several copies of the abstract store in it.
%
As it is, this semantics is much less efficient (but still more precise) than the previously proposed semantics because membership checks have to compare entire stores.
%
Checking equality is expensive because the stores within each state are large, and nearly every entry must be checked against every other due to high similarities amongst stores.
%%

%%
And, there is a better way.
%
Shivers' original work on $k$-CFA was susceptible to the same problem, and he suggested three complementary optimizations:
%
 (1) make the store global;
%
 (2) update the store imperatively; and
%
 (3) associate every change in the store with a version number -- its timestamp.
%
Then, put timestamps in states where previously there were stores.
%
Given two states, the analysis can now compare their stores just by comparing their timestamps -- a
constant-time operation.

@@

We lift $\dmachstep$ to accommodate for the asymmetry
in the input and output, and change the frontier-based semantics in
the following way:
{\small
\begin{align*}
(\mseentime, F, \msto,\Sigma,t) &\mathrel{\damachstep} (\mseentime \cup \mseentime', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(c',\msdiff) \mid \ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{c \mid (c,\_) \in I, (c,t') \notin \mseentime} \\
 \mseentime' &= \setof{(c,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}}
%%

%%
Here $\appendall$ combines change logs across all non-deterministic steps for a state to later be replayed.
%
The order the combination happens in doesn't matter, because join is associative and commutative.
%%

%%
\begin{lemma}
$\ttuple{c}{\msto} \dmachstep \ttuple{c'}{\msdiff}$ iff $\wn(c,\msto) \machstep \wn(c',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.
%%

%%
\begin{lemma}[$\changep$ means change]
Let $\replay(\msdiff,\msto) = \ttuple{\msto'}{\changep}$. $\msto' \neq \msto$ iff $\changep$.
\end{lemma}
By induction on $\msdiff$.
%%

%%
\begin{theorem}
$\damachstep$ is a complete abstraction of $\widehat{\machstep}^T$.
\end{theorem}
Follows from previous lemma and that join is associative and commutative.

@@

Tracing the execution of the analysis reveals an immediate shortcoming: there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)}, where {\tt f}, {\tt x} and {\tt y} each have several values each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective application positions.
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.2]{fanoutdot2}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism until and unless it is needed in {\it strict contexts} (such as the guard of an {\tt if}, a called procedure, or a numerical primitive application).
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.2]{lazydot}
\end{center}
%%

%%
This shift does not change the concrete semantics of the language to be lazy.
%
Rather, it abstracts over transitions that the original non-deterministic semantics steps through.
%
We say the abstraction is \emph{lazy} because it delays splitting on the values in an address until they are \emph{needed} in the semantics.
%
It does not change the execution order that leads to the values that are stored in the address.

@@

\renewcommand{\ext}{\mathit{ext}}
{\small
\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\msto,\mval) &= \setof{\mval}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\mkaddr}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\mval,\msto}
&\lmachstep\;
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\mkaddr),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take.
%
The middle rule uses $\force$ only to combine with values in the store - it does not introduce needless non-determinism.
%%

%%
\spchoice{
\noindent
We have two choices for how to implement lazy non-determinism.

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the baseline.
%
Consider a configuration where a reference to a variable and a binding of a variable will happen in one step, since store widening leads to stepping several states in one big ``step.''
%
With laziness, the reference will mean the original binding(s) of the variable \emph{or} the new binding, because the actual store lookup is delayed one step (i.e. laziness is administrative).

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the loss in precision by storing the result of the lookup in a value representing a delayed nondeterministic choice.
%
This is a more common choice in 0CFA implementations we have seen, but it interferes with the next optimization due to the invariant from store deltas we have that lookups must not depend on the change log.

% \paragraph{Our choice: option 1}
% The conditions for precision loss happen too rarely to warrant the
% added complexity to the next section's semantics to avoid such
% losses. Without abstract compilation, option 2 is the better
% choice.
}{}

@@

Thus far, we have made our optimizations in a purely functional manner.
%
For the final push for performance, we need to dip into the imperative.
%
In this section, we show an alternative representation of the store and seen set that are more space-efficient and are amenable to destructive updates by adhering to a history for each address.
%%

%%
The following transfer function has several components that can be destructively updated, and intermediate sets can be elided by adding to global sets.
%
In fact, the log of store deltas can be removed as well, by updating the store in-place, and on lookup, using the first value timestamped $\le$ the current timestamp.
%
We start with the purely functional view.
